---
title: "Distributional Consequences of the Clean Energy Transition"
author: "Natalie Smith"
format: 
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
theme: Litera
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

Use the following set of facts:

-   We are concerned about the environmental inequities that may arise between DAC and non DAC communities in California due to mining of lithium needed to produce batteries in electric vehicles (EVs). We therefore will model the California EV market with two consumer groups: DAC and non-DAC.

-   Initially, there is no intervention in the EV market.

-   The current EV price (without any intervention) is \$50,000 per vehicle.

-   The marginal cost of producing an EV is linear and has a price-intercept of 0. For the purposes of this exercise, you can assume that California is self-contained in supply and demand (you don’t need to worry about global EV or mining markets). This is of course unrealistic, but note there have been large efforts by the Biden Administration to create a domestic battery supply chain, making the tradeoffs analyzed here highly policy relevant.

```{r}
library(tidyverse)
```

1)  The data set HW3 data.csv provides price (in \$) and quantity (in number of EVs) estimates of demand per year for both DAC and non-DAC groups. Run linear regressions (with an intercept) to estimate the demand curves for DAC and non-DAC consumers. What are 2-3 reasons you can think that may explain the differences in demand you estimate?

```{r}
# Load the data
data <- read.csv("HW3_data.csv") %>% 
  janitor::clean_names()
```

```{r}
# Run linear regressions (with an intercept) to estimate the demand curves for DAC and non-DAC consumers
dac <- lm(price_dollars ~ q_dac, data = data)
non_dac <- lm(price_dollars ~ q_non_dac, data = data)
```

```{r}
#plot the linear regression for DAC
ggplot(data, aes(x = q_dac, y = price_dollars)) +
  geom_point() +
  labs(title = "DAC Demand Curve",
       x = "Quantity",
       y = "Price") +
  geom_abline(intercept = coef(dac)[1], slope = coef(dac)[2], color = "red")
```

```{r}
#plot the linear regression for non-DAC
ggplot(data, aes(x = q_non_dac, y = price_dollars)) +
  geom_point() +
  labs(title = "DAC Demand Curve",
       x = "Quantity",
       y = "Price") +
  geom_abline(intercept = coef(non_dac)[1], slope = coef(non_dac)[2], color = "red")
```

Reasons that may explain the differences in demand estimates:

-   Income: DAC communities might experience lower income levels compared to non-DAC communities, potentially limiting their capacity to afford electric vehicles.

-   Education: DAC communities may have lower levels of education compared to non-DAC communities, impacting their familiarity with EV technology and its advantages.

-   Availability of Charging Stations: DAC communities may encounter limited access to charging stations compared to non-DAC communities, potentially influencing their inclination towards EV adoption.

2)  One kilogram (kg) of lithium is estimated to cause \$300 dollars worth of damage in health costs to local communities living near mines. Each EV requires a lithium ion battery containing 63kg of lithium. Assuming there are no other externalities caused by EV production, what is the marginal externality cost per EV produced?

```{r}
# Define the marginal externality cost per EV produced
marginal_ext_cost <- 300 * 63
```

The marginal externality cost per EV produced is \$18,900.

3)  Describe the current equilibrium:

<!-- -->

(a) What is the aggregate annual demand curve for EVs?
(b) What is the supply curve for EVs?
(c) What is the “benefit” to consumers under the status quo?
(d) What is the “benefit” to producers under the status quo?
(e) What is the environmental cost under the status quo?
