---
title: "TBD"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(janitor)
library(here)
library(broom)
library(car)
library(pracma)
```

## 1.

The data set HW3 data.csv provides price (in \$) and quantity (in number of EVs) estimates of demand per year for both DAC and non-DAC groups. Run linear regressions (with an intercept) to estimate the demand curves for DAC and non-DAC consumers. What are 2-3 reasons you can think that may explain the differences in demand you estimate?

```{r}
# Load the data
data <- read.csv("HW3_data.csv") %>% 
  janitor::clean_names()
```

```{r}
# Run linear regressions (with an intercept) to estimate the demand curves for DAC and non-DAC consumers
dac_lm <- lm(price_dollars ~ q_dac, data = data)
dac_coeff <- dac_lm$coefficients

non_dac_lm <- lm(price_dollars ~ q_non_dac, data = data)
non_dac_coeff <- non_dac_lm$coefficients
```

```{r}
#pivot the data to have all dac and non-dac in one column
data_pivot <- data %>% 
  pivot_longer(cols = c(q_dac, q_non_dac), names_to = "group", values_to = "quantity")
```

```{r}
#plot the linear regression for both DAC and non-DAC
ggplot(data_pivot, aes(x = quantity, y = price_dollars)) +
  geom_point(aes(color = group)) +
  labs(title = "DAC and Non-DAC Supply & Demand Curves",
       x = "Quantity",
       y = "Price") +
  geom_abline(aes(intercept = coef(dac_lm)[1], slope = coef(dac_lm)[2]), color = "dodgerblue") +
  geom_abline(aes(intercept = coef(non_dac_lm)[1], slope = coef(non_dac_lm)[2]), color = "orangered1") +
  geom_hline(yintercept = 18900, color = "palegreen3") +  # marginal externality cost line
  geom_abline(intercept = 0, slope = 1, color = "orange") + # supply curve
  scale_color_manual(values = c("dodgerblue", "orangered1"), 
                     labels = c("DAC", "Non-DAC")) +
  # scale_linetype_manual(values = c("orange" = "solid", "palegreen3" = "solid"), 
  #                       labels = c("Supply", "Marginal Externality Cost")) +
  scale_y_continuous(limits = c(0, 180000)) +
  scale_x_continuous(limits = c(0, 80000)) +
  theme_bw()
```

Reasons that may explain the differences in demand estimates:

-   Income: DAC communities might experience lower income levels compared to non-DAC communities, impacting their willingness to pay.

-   Education: DAC communities may have lower levels of education compared to non-DAC communities, impacting their familiarity with EV technology.

-   Availability of Charging Stations: DAC communities may encounter limited access to charging stations compared to non-DAC communities, potentially influencing their inclination towards EV use.

## 2. Marginal Externality Cost:

One kilogram (kg) of lithium is estimated to cause \$300 dollars worth of damage in health costs to local communities living near mines. Each EV requires a lithium ion battery containing 63kg of lithium. Assuming there are no other externalities caused by EV production, what is the marginal externality cost per EV produced?

```{r}
# Define the marginal externality cost per EV produced
MEC <- 300 * 63
```

The marginal externality cost per EV produced is \$18,900.

## 3. Describe the current equilibrium:

### a. What is the aggregate annual demand curve for EVs?
```{r}
### Write demand functions

### P = a + bQ

dac_demand <- function(Q){
  P = 116854.84 - 6.61*Q
  return(P)
}

nondac_demand <- function(Q){
  P =  158034.28 - 2.73*Q
  return(P)
}

### Q = (P - a)/b

dac_demand_q <- function(P){
  Q = (116854.84 - P)/6.61
  return(Q)
}

nondac_demand_q <- function(P){
  Q = (158034.28 - P)/2.73
  return(Q)
}
```

**DAC Demand:** $P = 116,854.84 - 6.61*Q$

**Non-DAC Demand:** $P =  158,034.28 - 2.73*Q$

```{r}
### Aggregate demand curve
agg_demand <- function(P) {
  if (P > 116854.84) {
  Q = nondac_demand_q(P)
  }
  if(P <= 116854.84) {
   Q = dac_demand_q(P) + nondac_demand_q(P)
  }
  return(Q)
}
```

**Aggregate demand** $Q = -1.96Q + 148169$

###b. What is the supply curve for EVs?

EVs are currently at a price of \$50,000. We can plug this into our aggregate demand equation to find Q: 

```{r}
agg_demand(50000) 
### Q = 49687.19

50000/agg_demand(50000)
### 1.006296

supply <- function(Q) {
  P = 0 + 1.006296*Q
  return(P)
}
```
The supply curve for EVs is: $P = 1.006296Q$

```{r}
# ### P* and Q*
pstar <- 50000
qstar <- agg_demand(50000)
```


###c. What is the “benefit” to consumers under the status quo?
difference between the maximum price consumers are willing to pay ($50,000) and the actual price they pay.

```{r}
### find consumer surplus

cs_fun <- function(a, Q) {
  P = 1/2*(a - 50000)*Q
  return(P)
}

dac_cs <- cs_fun(116854.84, dac_demand_q(50000))
nondac_cs <- cs_fun(158034.28, nondac_demand_q(50000))
cs <- cs_fun(116854.84, dac_demand_q(50000)) + cs_fun(158034.28, nondac_demand_q(50000))

### DAC CS: 338,091,500
### Non-DAC CS: 2,137,620,083
### Total CS: 2,475,711,583

```

**Consumer Surplus:** $2,475,711,583$ 

###d. 
What is the “benefit” to producers under the status quo?

```{r}
### find producer surplus

ps_fun <- function(Q) {
  P = 1/2*50000*Q
  return(P)
}

ps_fun(agg_demand(50000)) 
### 1,242,179,825
```
**Producer Surplus:** $1,242,179,825$

###e. 
What is the environmental cost under the status quo?

```{r}
### EC = MEC* Q
ec <- qstar*MEC
# EC ### 939087948
```

**Environmental Cost:** $939,087,948$

```{r}
library(ggplot2)

ggplot(data = data_pivot, aes(x = quantity, y = price_dollars, color = group)) + 
  geom_point() + 
    labs(title = "DAC and Non-DAC Supply & Demand Curves",
       x = "Quantity",
       y = "Price") +
  stat_function(fun = function(Q) dac_demand(Q), color = "dodgerblue") +
  stat_function(fun = function(Q) nondac_demand(Q), color = "orangered1") + 
  stat_function(fun = function(Q) supply(Q), color = "darkgreen") +
  geom_hline(yintercept = 18900, color = "orange") +
  scale_color_manual(name = "Group",
                     values = c("dodgerblue", "orangered1"), 
                     labels = c("DAC", "Non-DAC")) +
  ylim(0, 190000) +
  xlim(0, 90000) +
  theme_minimal()

```

## 4. How is the current consumer benefit divided between DAC and non-DAC consumers?

```{r}
### DAC CS: 338,091,500
### Non-DAC CS: 2,137,620,083

cb_dac <- cs_dac/cs
cb_non_dac <- cs_non_dac/cs

```

Current consumer benefit divided between DAC and non-DAC consumers:
DAC: $13.67%$
Non-DAC: $86.33%$

### 5. 
Derive the optimal EV tax (in \$ per vehicle) to internalize the lithium mining externality. Noting that recent research has shown that DACs are far more likely to contain mining activity, assume that the mining externality is borne entirely by the DAC group. What would be the effects of this tax on:

###a. The amount of EVs produced and consumed

Because MEC is constant, optimal tax = $18,900/EV$

```{r}
msc <- function(Q) {
  P = MEC + 1.006296*Q
  return(P)
}

msc_q <- function(P) {
  Q = (P - MEC)/1.006296
  return(Q)
}

agg_demand_p <- function(Q){
  P <- -1.96*Q + 148169
  return(P)
}

F3 <- function(Q){
  z <- agg_demand_p(Q) - msc(Q)
  return(z)
}

fzero(F3, c(0,80000))$x

qstar_soc <- 43579.27

#plug in p-star to find the price
p_star_soc <- msc(qstar_soc)
p_star_soc <- 62430.86

```

The amount of EVs produced and consuned would decrease from 50,000 to 43,892.08 units. 

###b. The price of EVs

The price of each EV would increase from 50,000 to 62,430.86

###c. Overall welfare of non-DAC consumers

```{r}
cs_non_dac_welfare <- 1/2*(158034.28 - p_star_soc)*nondac_demand_q(p_star_soc)
###  1,673,995,223
# cs_non_dac_welfare
```
#REWRITE w updated numbers :Benefit to non-DAC Consumers w/tax = 1,454,620,139 <br> This is a decrease of 682506553 from a non taxed CS of 2,137,620,083

###d. Overall welfare of DAC consumers

```{r}
cs_dac_welfare <- 1/2*(116854.84 - p_star_soc)*dac_demand_q(p_star_soc)
### 224,052,163
cs_dac_welfare 
```

#REWRITE THIS with updated numbers: Benefit to DAC Consumers w/tax = 173,953,607. This is a decrease of 164,137,893 from the non taxed CS of 338,091,500. The overall welfare for DAC will be their benefit after the tax is implemented 

###e. EV producers
```{r}
# original ps = 1,241,954,750

ps_tax <- 1/2 * (qstar_soc) * (p_star_soc-18900)
# 941,539,757

# ps_tax

```

EV producers would experience a decrease in producer surplus from 1,242,179,825 to 941,539,757.


###f. Total environmental damage

```{r}
### Environmental cost = MEC ($18,900) * Qagg(50,000)
env_cost_tax <- 18900 * qstar_soc
# env_cost_tax
# = 817,585,650

#enviro cost (no tax) = 939087948

envirodamage <- ec - env_cost_tax 
envirodamage

```

Total environmental damage would decrease by 184,887,039, from 939,087,948 to 754,200,909

###g. Total tax revenue generated

The same as the total environmental damage, $817,585,650.

## 6.
Now, assume that all revenue from the EV tax will be redistributed to the consumers in proportion to their pre-tax consumption of EVs. For example, if 80% of the EVs were consumed by non-DAC consumers, then they get 80% of the tax revenue. Calculate the effects of this EV tax on:
### a.Overall welfare of non-DAC consumers
```{r}

```

### b.Overall welfare of DAC consumers
### c.EV producers

## 7.
 Based on your findings, use 1-3 sentences to answer the following (qualitatively only):

### a.
Are non-DAC consumers likely to support the EV tax? How would the redistribution regime in question #6 affect their support for the tax?


### b.
Are DAC consumers likely to support the EV tax? How would the redistribution regime in question #6 affect their support for the tax?
